---
title: "ps5-rmarkdown"
author: "Maggie"
date: "2023-02-18"
output: html_document
---

```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("data/gapminder.csv.bz2") 
```

```{r}
gapminder %>% 
  dim()
```
```{r}
gapminder %>% 
  sample_n(5)
```

```{r}
gapminder %>% 
  summarize(iso2sum = n_distinct(iso2)) %>% 
  pull(iso2sum)

gapminder %>% 
  summarize(iso3sum = n_distinct(iso3)) %>% 
  pull(iso3sum)

gapminder %>% 
  summarize(n = n_distinct(name)) %>% 
  pull(n)
```

iso2 has more names than iso3

```{r}
gapminder %>% 
  count(iso2, name,)
```

```{r}
gapminder %>% 
  summarize(mintime = min(time, na.rm = TRUE),
            maxtime = max(time, na.rm = TRUE)) %>% 
  pull(mintime, maxtime)
```
The max year is 2019, and the min year is 1960. 



```{r}
gapminder %>% 
  select(co2) %>% 
  summarize(sum(is.na(.)))
```
```{r}
gapminder %>% 
  select(co2_PC) %>% 
  summarize(sum(is.na(.)))
```
co2_PC(2661) has more missing data than co2(2658).

```{r}
carbonemission <- gapminder %>% 
  group_by(co2) %>% 
  filter(name == "United States of America" |name == "China" | name == "India" |
           name == "Japan" | name =="Italy") 

ggplot(carbonemission) +
  geom_col(aes(x = name, y = co2))
```
Of these 5 countires, the US has the most carbon emissions, then followed by China. Italy has the least CO2 emissions in 
comparison to the rest. 

```{r}
carbonemissionpercap <- gapminder %>% 
  group_by(co2_PC) %>% 
  filter(name == "United States of America" |name == "China" | name == "India" |
           name == "Japan" | name =="Italy") 

ggplot(carbonemissionpercap) +
  geom_col(aes(x = name, y = co2_PC))
```
The **United states** still has the most carbon emissions per capita, meaning that the average person in America 
will emit more CO2 than the average person of any of the remaining countries. India's CO2 emission per capita 
is the lowest because they don't emit a lot of CO2 in comparison to the US and China and they have a much higher 
population among most countries. 

```{r}
gapminder %>%
  group_by(region) %>% 
  summarise(n = mean(co2_PC, na.rm = TRUE))
  
```
The average per capita CO2 emissions _over time_ indicates that the region, Europe has the highest per capita emission (
if we omit the NA row). Africa is the has the smallest average per capita emission. 

```{r}
percap <- gapminder %>%
  group_by(region) %>% 
  filter(time == 1960 | 2016) %>% 
  summarise(n = mean(co2_PC, na.rm = TRUE))

ggplot(percap) +
  geom_bar(aes(x = region), color = time)
```




